Nengo: A Python tool for building large-scale functional brain models

### 2013-12-04 rebuttal.txt

Dear Editor,

Please find a revised version of our Methods paper "Nengo: A Python
tool for building large-scale functional brain models" attached for
your consideration for publication in Frontiers in Neuroinformatics.

Sincerely,

Trevor Bekolay, James Bergstra, Eric Hunsberger, Travis DeWolf,
Terrence C. Stewart, Daniel Rasmussen, Xuan Choo, Aaron Voelker, and
Chris Eliasmith

Reviewer 1
==========

Expert Authority
----------------

>> Are relevant methods of other authors mentioned? If not, please
>> specify the methods the author should incorporate.

> Emergent should be mentioned. It is not based on spiking simulators,
> but has a track record in cognitive modelling.

We have now cited Emergent in the introduction alongside ACT-R in the
introduction.

> MIIND could be mentioned (Disclaimer: the reviewer is involved with
> MIIND) It models population dynamics, admittedly not based on
> spiking neurons, but again its aim is large-scale cognitive models
> and shares some design ideas (see paper in Neural Networks). Please
> judge for yourself.

We have revised Section 7.1, which now includes a list of tools
that simulate cognitive phenomena, including MIIND.

The revised text reads:

"There are many other neural simulators dedicated to building
large-scale neural models (Eppler et al., 2008; Goodman and Brette,
2009; Hines et al., 2009), and many tools for simulating cognitive
phenomena with various levels of biologically plausibility (Cooper and
Fox, 1998; Sun, 2001; Anderson et al., 2004; Franklin et al., 2007;
Aisa et al., 2008; de Kamps et al., 2008; Laird, 2012). However, Nengo
is unique in being built on a theoretical framework that has enabled a
cognitive architecture (the Semantic Pointer Architecture) that
maintains a high level of biological plausibility, and has been
validated through the Spaun model and other past work."

Other Comments
--------------

>> Please add any further comments you have regarding this manuscript.

> It looks like an interesting tool, and I think at this level
> existing simulators can do with a bit of competition. But the
> description of the tool leaves much to be desired. In particular it
> doesn't make clear enough what its advantages over NEST or BRIAN
> are. The discussion on scaling is not entirely convincing because in
> part, it is hard to understand how to compare with those simulators.

We have expanded Section 7.1 to compare to projects that have similar
design goals and present a similar interface (e.g., PyNN) and between
the underlying neural simulators (e.g., NEST, Brian).

The revised section on neural simulators reads:

"On the neural simulator side, we have shown that both Nengo's
reference simulator and OpenCL simulator are able to simulate two
benchmark models much faster than Brian, NEST and NEURON (see
Figure~\ref{fig:benchmarks}). This is, in part, because Nengo stores
the factors of the connection weight matrix, rather than storing the
entire matrix. However, these simulators are able to simulate many
detailed neuron models and learning rules, and have access to a wealth
of existing neuron models and learning rules. Because Nengo 2.0 is in
an earlier development stage, many of these detailed neuron models and
learning rules remain to be added. Neural simulators like Brian, NEST,
and NEURON are therefore currently better suited for simulating a
wider range of single cell models, while Nengo is designed for large
networks of simple neural models that are connected together according
to the principles of the Neural Engineering Framework, and can
simulate these types of models efficiently.

"One key difference between Nengo's simulators and traditional neural
simulators is the target platform. While NEST and NEURON can be run on
commodity hardware, networks of modest size are typically simulated in
high-performance computing environments by using the Message-Passing
Interface (MPI). Nengo enables large-scale neural simulation on
commodity hardware, allowing researchers and hobbyists without access
to large computing clusters the ability to experiment with theoretical
neuroscience as is currently practiced in cutting edge research. In
particular, GPUs are a powerful, low-cost computing resource that are
available in most modern workstations. The OpenCL simulator makes full
use of GPUs, while the previously discussed simulators currently do
not."

> Where is it available? What do I need in terms of 3rd party software?

Thanks for pointing out that we omitted these important details in the
original manuscript. We have added a link to the release and source
versions of Nengo in a footnote on Page 2 of the revised manuscript
that reads:

"Nengo is available for download through the Python Package Index at
https://pypi.python.org/pypi/nengo. The full source is available at
https://github.com/ctn-waterloo/nengo."

We have also added the following paragraph to the introduction in
order to clarify Nengo's 3rd party dependencies.

"Nengo 2.0 has been rewritten from scratch in Python, leveraging NumPy
(Oliphant, 2007) for manipulating large amounts of data. While NumPy
is its only dependency, Nengo contains optional extensions for
plotting if Matplotlib is available (Hunter, 2007) and for interactive
exploration if IPython 63 available (Perez and Granger, 2007). Since
Nengo only depends on one third-party library, it is easy to integrate
Nengo models in arbitrary CPython programs, opening up possibilities
for using neurally implemented algorithms in web services, games, and
other applications."

> The NEF should be explained in more detail. The paper should be more
> self-contained. It is not entirely clear how the simulations are
> implemented. I think an ensemble of, for example, LIF neurons is
> simulated, and the dynamics and input that they receive AS A
> POPULATION is presented. That is interesting, but I'm not entirely
> sure how this is done. What happens in the Lorenz attractor? Do all
> neurons receive feedback according to the differential equations in
> the script, i.e. as a current? Or as spike trains modulated by this
> system? What is 'input' anyway? This is not described very well at
> all. I would walk the reader through a simple simulation more
> explicitly.

In response to your direct question, all of the neurons in the Lorenz
attractor receive input in the form of current; that current is
determined by a summation of the spiking activities of afferent
neurons weighted by connection weights that are determined using the
NEF's transformation principle.

These details are now included in Appendix 1, "Neural Engineering
Framework Details", which we have added to make this paper more
self-contained. This appendix gives the mathematical details behind
the principles discussed in Section 2. (The text of the Appendix is
too long to include in the interactive review forum.)

We have also described the example scripts in more detail in
sections 5.1, 5.2, and 5.3.

In regard to walking the reader through a simple simulation, we hope
that the addition of the NEF appendix will make the existing Figure 2
a sufficiently detailed account of the first two timesteps of a simple
simulation.

> Having said that, the paper contains far too many irrelevant design
> details. The paragraph on JAVANENGO is bizarre. Do you really want
> to burden you readers with the fact that its class interfaces are
> overloaded? If you want to promote the new Python tool, there is no
> reason to blast the Java version. I recommend dropping the entire
> JAVANENGO section, if the Python version is what you want people to
> use.

Thanks for pointing this out. As the designers of the new version, we
have had many discussions about whether we should improve the old
version or start from scratch; some of that discussion appeared in the
paper, and upon a second look, we wholeheartedly agree that the
mentions of JavaNengo are out of place.

We have removed that section completely in the revised manuscript.
Where mentions of the previous version of Nengo were unavoidable, we
have referred to JavaNengo as Nengo 1.4, and the new version as Nengo
2.0 rather than PyNengo.

> I would mention design considerations only when the relate to use
> cases, i.e. explain why certain scaling properties are achieved. The
> paper contains far too many design considerations, it should focus
> on use cases. Scaling is also an issue of use.
>
> This question is important in order to judge how similar the
> simulator is from other tools such as NEST and BRIAN. Can you set up
> simulations in the two of those that are similar to NENGO and THEN
> compare them? Also bear in mind that NEST has an MPI backbone.

We wanted this paper to focus on the design of Nengo 2.0, and so our
intention in saying that the implemented simulators scale well is
primarily to indicate that the design enables the implementation of
simulators that scale well. It is difficult to come up with use cases
that can be adequately explained in this paper; in a sense, the papers
that are cited in the introduction are those use cases, as they were
all implemented with Nengo 1.4 (Conklin and Eliasmith, 2005; Singh and
Eliasmith, 2006; Choo and Eliasmith, 2010; Rasmussen and Eliasmith,
2014; DeWolf and Eliasmith, 2011; Stewart et al., 2012; Eliasmith et
al., 2012). We believe that the examples discussed in Section 5 are of
sufficient length to be instructive.

Setting up simulations first in Brain or NEST (or PyNN) and then
mapping those to Nengo models would not be very instructive. The point
of Nengo is to build neural models based on the principles of the NEF;
models built in those other tools would not use these principles, and
would result in models containing many ensembles whose underlying
neural populations would be connected directly with a defined weight
matrix. We do make this possible for certain important cases (e.g.,
inhibitory gating signals), but these cases are not the norm when
using the principles of the NEF. We hope that the focus of the paper
is clearer given our additions in the introduction and in Section 7.1.

We agree that it would be an interesting comparison to simulate two
neural populations connected directly with a defined weight matrix.
However, that comparison would be more apt in a paper that gives more
implementation details of the simulators being compared. We have not
provided those details here, and have instead focused on the novel
contributions of Nengo.

In sum, the difference between Nengo and other neural simulators is
primarily that Nengo is designed to define and simulate neural models
created with the principles of the NEF. The updated introduction and
Section 7.1 (copied above) highlights this difference.

> Can you do other neural models than LIF? Again this harks back to
> the criticism that the philosophy of NEF is well described, and
> interesting, but its implementation is not.

Currently in the main development branch of Nengo 2.0, we have only
implemented the LIF model, and a rate approximation of the LIF model.
Nengo 1.4 contained implementations of LIF, an adaptive LIF model
(Koch & Schutter, 1999), and the well-known Hodgkin-Huxley model, so
we know that it is possible to use other neuron models in NEF-based
networks. Additionally, we have already completed work done on another
simulator internal to the lab that uses a neuron model emulating a
piece of neuromorphic hardware.

The added appendix describing the NEF makes explicit how the neuron
model is used. Additionally, we have added the following text to
Section 2:

"Importantly, tuning curves can be determined for any type of neuron,
and therefore the encoding process (and the NEF as a whole) is not
dependent on any particular neuron model."

> Two million spiking neurons is not that much. Edelman and Tononi did
> that 15 years ago. Izhekevich blasts you out of the water. Moreover,
> it is a meaningless quantity, it depends on what you do. 2.5 million
> compartmental neurons is still impressive, 2.5 point model neurons is
> not. In our lab have models with hundreds of thousands of rate-based
> populations, we could model each population by a thousand neurons if
> we wanted to. Does that make us record holder? I'd be a bit more
> careful with such claims, or qualify them better.

We agree that the number of neurons is a meaningless quantity, and
believe that function is the more important metric. That is why our
focus is well-exemplified by Spaun, the largest functional brain model
(which has been previously stated as such in Eliasmith & Trujillo,
2013), where functional in this case refers to the cognitive functions
that Spaun performs.

> In my opinion, this paper cannot be read without going to the original
> paper explaining NEF. Given the brevity of this paper, I think that's
> not good enough. There is ample space for a better explanation of how
> it works and how that is different from more conventional simulators.

Thank you for pointing out these weaknesses. We hope that the addition
of an appendix describing the NEF, an expanded Section 7.1 comparing
Nengo to conventional simulators, and other improvements throughout
the revised manuscript have addressed your concerns and have made this
paper self-contained. In particular, we hope that we have made it
clear that Nengo's strength lies in the fact that it is based on the
principles of the NEF; we do not wish to claim that Nengo is a better
general-purpose neural modeling tool than PyNN, or a better
general-purpose simulator than Brain or NEST. However, we do wish to
claim that Nengo is a better modeling tool and simulator for NEF-style
networks, and that NEF-style networks are less computationally
intensive to simulate, and therefore are feasible to run on commodity
hardware.

Reviewer 2
==========

Expert Authority
----------------

>> Are relevant methods of other authors mentioned? If not, please
>> specify the methods the author should incorporate.

> Authors could perhaps mention the Topographica
> (http://topographica.org/index.html) simulation framework which is
> also implemented in python and offers a high-level model
> specification environment. It also offers a lot of analysis and
> visualization resources and a GUI, which could serve as inspiration
> for some of the future work mentioned in the manuscript.

Thank you for bringing Topographica to our attention! While we have
looked at it some time ago, the project seems to have grown much
lately, and we will investigate it more soon. For now, we have cited
it in Section 7.1, as it seems to have similar goals as Nengo and
PyNN.

Article Length
--------------

>> A Method Article should not exceed 12,000 words. Should any part of
>> the article be shortened? If yes, please specify which part should be
>> shortened.

> I found the introduction unnecessarily lengthy with rather small
> proportion actually discussing the main topic (pyNENGO). Especially
> the JavaNENGO section could in my opinion be omitted or at least
> significantly shortened. Also I found the first part of the
> introduction discussing NEF unnecessarily long - this paper should
> motivate pynengo not NEF.

Thanks for pointing this out. As the designers of the new version, we
have had many discussions about whether we should improve the old
version or start from scratch; some of that discussion appeared in the
paper, and upon a second look, we wholeheartedly agree that the
mentions of JavaNengo are out of place.

We have removed that section completely in the revised manuscript.
Where mentions of the previous version of Nengo were unavoidable, we
have referred to JavaNengo as Nengo 1.4, and the new version as Nengo
2.0 rather than PyNengo.

We have also removed some sections of the introduction motivating the
NEF, and have replaced them with motivations for Nengo. Specifically,
the paragraph:

"It has often been said that neuroscience is data-rich and theory-poor
(Churchland and Sejnowski, 1992). We believe that the modeling success
of large functional networks such as Spaun suggests that the NEF may
be able to fill the theoretical void. In order to continue to evaluate
the NEF as a theory of neural computation by building larger and more
complicated models, we have begun the next generation of Nengo
development with a focus on speed, extensibility, and simplicity."

has been removed, and replace with the following text:

"The transformation principle of the NEF proposes that the connection
weight matrix between two neural populations can compute a nonlinear
function, and can be factored into two significantly smaller matrices.
By using these factors instead of full connection weight matrices,
NEF-designed models are more computationally efficient, which allows
Nengo to run large-scale neural models on low-cost commodity
hardware."

Other Comments
--------------

> Section 1.2 alludes to the possibility of exporting model from pynengo
> to PyNN. Is this a future work, or is there already some
> infrastructure in place for this? Please clarify.

The ability to export a model to PyNN is a future work. We have
removed that confusing passage from Section 1.2 in the revised
manuscript, but that functionality is possible and we hope to
implement it soon.

We have already have two work-in-progress simulators allowing for
Nengo 2.0 models to be run on two pieces of neuromorphic hardware. We
have added a mention of these, copied below, to Section 7.2 in order
to show that multiple backend simulators are possible (including a
future PyNN backend):

"We are also developing two simulators that will take the same NEF
model description as the existing simulators, but will target two
pieces of neuromorphic hardware to achieve greater speed and power
efficiency than the OpenCL simulator."

> In section 3.2 authors discuss parameters, some of which can be set on
> ensembles and some directly on neural populations. It isn't very
> intuitive why some belong to one or the other (i.e. why maximum firing
> rate is a parameter of ensemble rather than the population). I feel it
> could be instructive to discuss a bit more in detail the separation of
> the high level NEF concepts from the low-level spiking neuron concepts
> in pynengo.

The intuition as to why some parameters are ensemble-level and some
are neural population-level relies on a mathematical description of
how we compute the current that we inject into each of the neuron in
the neural population. Any parameter that is part of that equation is
ensemble-level, and any parameter that is not is stored on the neural
population. We have now included these equations in an appendix that
gives a mathematical description of the NEF; specifically, Equation
(1) describes the current that is injected into each neuron model. The
parameters of this equation are stored on the ensemble, whereas the
parameters that are used in the neural nonlinearity would be stored in
the neural population object.

In the revised manuscript we have also removed the mention of the
maximum firing rate in Section 3.1 (formerly 3.2), as the encoding
weight is more intuitively a part of the NEF's representation scheme
specifically and not a property of a neuron.

> I was intrigued by the mention of automatic parameter setting based on
> neurobiological constraints. Could authors discuss this issue further
> in the manuscript?

The description "automatic parameter setting" may be too generous due
to our choice of wording. Specifically, in our original manuscript we
wrote in Section 3.2 (now 3.1):

"If these attributes are not set, Nengo attempts to maintain
neurobiological constraints by selecting neural parameters from
distributions consistent with neocortical pyramidal cells."

However, it is more accurate to say that, if parameters are unset,
Nengo will use defaults that correspond to neocortical pyramidal
cells. The text in the updated manuscript now reads:

"If these attributes are not set, Nengo attempts to maintain
neurobiological constraints by using default parameters consistent
with neocortical pyramidal cells."

We do wish to do some kind of automatic parameters setting in the
future, both in terms of neural parameters and connection parameters.
In the newly added NEF appendix, we also mention how connection
parameters can correspond to experimental data:

"The decoding process is depicted in Figure 1B, where the optimal
linear decoders have been found and used for eight neurons. Notably,
this kind of temporal decoding requires an assumption about the nature
of the temporal filter being used. Here we assume that post-synaptic
currents are such filters, and set the time constants to reflect the
kind of neurotransmitter receptors in the connection (e.g., AMPA
receptors have short time constants, ~10 ms, and NMDA receptors have
longer time constants, ~50 ms)."

> It isn't very clear from the manuscript what Network is. Is it really
> just a simplest form of a container as would appear from the
> description or does it actually offer some distinct functionality?

A Network really is just a simple container, which is used for
organizing large models. Our experience building large models has
shown that this kind of hierarchical organization becomes increasingly
useful as the size of the model increases, and we include Networks for
that reason. The text in Section 3.5 has been revised to explicitly
state that networks are only containers, and can be used for
hierarchical grouping.

> The first paragraph of section 4 implies a simulator dependent
> results. If this is the case, I think it warrants a more detailed
> discussion.

Thanks for pointing this out. We have revised this paragraph to remove
the simulator dependent language, and have moved that information to
Section 4.1, which collects together information specific to the
reference simulator.

The revised first paragraph of Section 4 is:

"Decoupling model creation and simulation has been done previously by
PyNN (Davison et al., 2008). In PyNN, the same Python script can be
used to run a model on four different simulators. Nengo follows this
programming model by decoupling neural model creation and simulation,
which enables Nengo simulators to allocate memory and schedule
computations in the most efficient ways possible. Simulators are given
a Model as an argument; this Model is a static symbolic description.
The simulator can take the model description and build whatever data
structures are best suited to implement the simulation."

> The benchmark results are impressive, but it should be noted that
> most spiking simulators these days are used in HPC environments, and
> thus parallelization comparisons in multi-processor environments
> would be more relevant. If pynengo does not offer this
> functionality, this should be acknowledged and perhaps future plans
> in this direction should be discussed.

### 2013-12-04 nengo.tex
Manuscript revised based on reviewer comments
- Tried to address all reviewer concerns
- Also included responses to reviewer comments in rebuttal.txt

\begin{abstract}
  Neuroscience currently lacks
  a comprehensive theory of
  how cognitive processes can be
  implemented in a biological substrate.
  The Neural Engineering Framework (NEF)
  proposes one such theory,
  but has not yet gathered
  significant empirical support,
  partly due to the technical
  challenge of building and simulating
  large-scale models with the NEF.
  Nengo is a software tool
  that can be used to build and simulate
  large-scale models based on the NEF;
  currently, it is the primary resource
  for both teaching how the NEF
  is used,
  and for doing research
  that generates specific
  NEF models
  to explain experimental data.
  A previous version of Nengo
  implemented in Java was used
  to create Spaun,
  the world's largest functional
  brain model \citep{eliasmith2012}.
  Simulating Spaun
  highlighted limitations
  in Nengo's ability to
  support model construction with simple syntax,
  to simulate large models quickly,
  and to collect large amounts of data
  for subsequent analysis.
  This paper describes a new version of Nengo
  implemented in Python that overcomes these limitations.
  It uses simple and extendable syntax,
  simulates a benchmark model on the scale of Spaun
  50 times faster than the previous implementation,
  and has a flexible mechanism
  for collecting simulation results.

@@

Modeling the human brain
is one of the greatest
scientific challenges of our time.
Computational neuroscience
has made significant advancements
from simulating low-level biological parts in great detail,
to solving high-level problems that humans find difficult;
however, we still lack a mathematical account of
how biological components
implement cognitive functions such as sensory processing,
memory formation, reasoning, and motor control.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological synapses \citep{markram2006}.
However, cognition has not yet emerged
from data-driven large scale models,
and there are good reasons to think
that cognition may never emerge \citep{trujillo-inpress}.
At the other end of the spectrum,
cognitive architectures \citep{anderson2004}
and machine learning approaches \citep{hinton2006}
have solved high-dimensional statistical problems,
but do so without respecting biological constraints.

Nengo is a neural simulator
based on a theoretical framework proposed
by \citet{eliasmith2003}
called the Neural Engineering Framework
(NEF).
The NEF is a large-scale modeling approach
that can leverage single neuron models
to build neural networks with
demonstrable cognitive abilities.
Nengo and the NEF has been used to build
increasingly sophisticated neural subsystems
for the last decade
(e.g., path integration [\citealp{conklin2005}],
working memory [\citealp{singh2006}],
list memory [\citealp{choo2010}],
inductive reasoning [\citealp{rasmussen2011}],
motor control [\citealp{dewolf2011}],
decision making [\citealp{stewart2012}])
culminating recently with Spaun,
currently the world's
largest functional brain model \citep{eliasmith2012}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
If asked to recall the memorized list,
Spaun generates motor commands for a simulated arm,
writing out the digits $1234$.
While the tasks that Spaun performs are diverse,
all of the tasks use a common set of
functional cortical and subcortical components.
Each functional component corresponds
to a brain area that has been hypothesized
to perform those functions
in the neuroscientific literature.

The NEF provides principles
to guide the construction of
a neural model
that incorporates anatomical constraints,
functional objectives, and
dynamical systems or control theory.
Constructing models from this starting point,
rather than from single cell electrophysiology
and connectivity statistics alone,
produces simulated data that explains
and predicts a wide variety of experimental results.
Single cell activity \citep{stewart2012},
response timing \citep{stewart2009a},
and behavioral errors \citep{choo2010}
of NEF-designed models match
psychological and physiological findings
without being built specifically into the design.
These results are a consequence of
the need to satisfy functional objectives
within anatomical and neurobiological constraints.

It has often been said that neuroscience
is data-rich and theory-poor \citep{churchland1992}.
We believe that the modeling success of
large functional networks such as Spaun
suggests that the NEF
may be able to fill the theoretical void.
However, in simulating this large model,
the Java-powered Nengo simulator
was limited in terms of memory consumption,
simulation speed, and extensibility.
In order to continue to evaluate the NEF
as a theory of neural computation
by building larger and more complicated models,
we have begun the next generation
of Nengo development with a focus on speed,
extensibility, and simplicity.

This paper introduces a new version
of Nengo (PyNengo) written in Python.
The remainder of this introduction outlines
the design issues in the previous version (JavaNengo)
that motivated the reimplementation.
PyNengo is not currently meant
to completely replace JavaNengo;
JavaNengo remains a capable visual NEF
simulator and a valuable teaching tool
due to its graphical interface.
PyNengo's focus is on speed
and a simplified internal design
to support ongoing research in
large adaptive functional models.
The remainder of the paper outlines PyNengo's design,
scripting interface, and performance relative to
JavaNengo and alternatives such as Brian and NEST.

\subsection{JavaNengo limitations}

\begin{table}[!t]
\processtable{Tools bearing the name Nengo.\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{Code repositories avilable at \url{http://github.com/ctn-waterloo/}.}
\end{table}

JavaNengo is a suite of tools composed of
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(as detailed in Table~\ref{tab:nengo-vers}).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of several
methods for creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
For example,
the most often used class for making
NEF models, \texttt{NEFEnsemble},
extends from five ancestor classes
and implements seven interfaces.
Adding support for synaptic plasticity
in \texttt{NEFEnsemble}
involved adding or significantly modifying
25 Java classes \citep{bekolay2010}.

The JavaNengo GUI provides
a drag-and-drop interface in which
NEF models can be created.
However, as it was developed
before the scripting interface,
the GUI manipulates simulator objects directly,
rather than constructing a script
that manipulates those simulator objects.
Typically, modellers start by following
tutorials that use the GUI interface,
and transition to the scripting interface
in order to implement more complex models;
since the GUI modifies simulator objects directly,
modelling concepts learned in the GUI
have to be relearned in the scripting interface.

The JavaNengo scripting interface,
described in \citet{stewart2009},
provides a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
allowing models to be described
by short scripts.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., to add new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As there are few neural modellers
that are proficient with both Java and Python syntax,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.
Additionally, Jython is always several versions
behind CPython, making it impossible
to use new language features
like dictionary comprehensions.
Jython also cannot access libraries
that contain any C code,
including all scientific libraries
that use NumPy.

Finally, perhaps the most significant limitation
of JavaNengo is its performance.
While Java has good support for threads,
it lacks native high-performance numerical libraries
for doing mathematical operations on large matrices.
Additionally, it is difficult to interface with
hardware such as graphical processing units (GPUs),
which are well-suited to the computations
required in neural simulations.

Addressing all of these issues
in the Java codebase would require
enough extensive changes
that we opted instead
to write a new version of Nengo in Python.
JavaNengo's Jython scripting interface
already used Python syntax,
and since Python is widely used
for neural simulation and data analysis,
we targeted the CPython interpreter
as the platform for the new version of Nengo.

\subsection{PyNengo}

PyNengo is a Python package for
defining and simulating
neural models using the NEF.
Its interfaces are based on NumPy,
so it integrates seamlessly
with other scientific Python tools
for interactive exploration \citep{perez2007},
scientific computation \citep{oliphant2007},
and plotting \citep{hunter2007}.
Arbitrary CPython programs
can use Nengo models,
opening up possibilities
for using neurally implemented algorithms
in web services, games, and other applications.

PyNengo has a simple object model,
which makes it easy to
document, test, and modify.
These objects are accessible
through a scripting interface
similar to JavaNengo's \citep{stewart2009},
with improvements
for easier incorporation of user-defined networks,
neuron models and learning rules.
Critically, and unlike JavaNengo,
PyNengo's scripting interface
contains functions to record simulated data
and access that data after simulation.

Model creation and simulation are decoupled,
allowing for models to be exported
to other simulation packages
(e.g., those accessible by PyNN [\citealp{davison2008}])
or for faster simulators to be used
as drop-in replacements for PyNengo's
platform-independent reference simulator.
To date, we have implemented one other simulator
that uses PyOpenCL to take advantage
of a GPU or multicore CPU.

PyNengo does not yet
replace all of JavaNengo's functionality.
While it can simulate most
models that JavaNengo can simulate,
there is currently no graphical interface for PyNengo.
Model components must be created
using Python scripts rather than
the drag-and-drop interface presented by JavaNengo.
Some of the means of interacting
with a running model in JavaNengo
can be provided using packages
such as IPython and Matplotlib,
but such tools have not been our focus thus far.
It is possible to build a graphical program
like JavaNengo's GUI on top of PyNengo's object model,
but that remains as future work.

To summarize, PyNengo overcomes
JavaNengo's limitations.
It can simulate most models
currently simulated by JavaNengo
using a well-tested codebase of
$\sim$5000 lines of code,
compared to JavaNengo's $\sim$45,000
(which does not include the JavaNengo GUI).
A new internal representation of neural models
allows the OpenCL simulator to simulate
large models on the scale of Spaun
hundreds of times faster than JavaNengo
using inexpensive commodity hardware.
Compatibility with CPython makes scientific Python software
available for preparing and analyzing data.
In all, PyNengo provides a platform for
simulating larger and more complex models than Spaun,
and can therefore further test the NEF
as a theory of neural computation.

\section{Neural Engineering Framework (NEF)}

@@

In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how likely
that neuron is to respond to a given input signal.
The tuning curve is a function of the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the \textit{encoding weight}
(the direction in the input vector space
that causes the neuron to be the most active).

>>>>

\begin{abstract}
  Neuroscience currently lacks
  a comprehensive theory of
  how cognitive processes can be
  implemented in a biological substrate.
  The Neural Engineering Framework (NEF)
  proposes one such theory,
  but has not yet gathered
  significant empirical support,
  partly due to the technical
  challenge of building and simulating
  large-scale models with the NEF.
  Nengo is a software tool
  that can be used to build and simulate
  large-scale models based on the NEF;
  currently, it is the primary resource
  for both teaching how the NEF
  is used,
  and for doing research
  that generates specific
  NEF models
  to explain experimental data.
  Nengo 1.4, which was implemented in Java,
  was used to create Spaun,
  the world's largest functional
  brain model \citep{eliasmith2012}.
  Simulating Spaun
  highlighted limitations
  in Nengo 1.4's ability to
  support model construction with simple syntax,
  to simulate large models quickly,
  and to collect large amounts of data
  for subsequent analysis.
  This paper describes Nengo 2.0,
  which is implemented in Python
  and overcomes these limitations.
  It uses simple and extendable syntax,
  simulates a benchmark model on the scale of Spaun
  50 times faster than Nengo 1.4,
  and has a flexible mechanism
  for collecting simulation results.

@@

Modeling the human brain
is one of the greatest
scientific challenges of our time.
Computational neuroscience
has made significant advancements
from simulating low-level biological parts in great detail,
to solving high-level problems that humans find difficult;
however, we still lack a mathematical account of
how biological components
implement cognitive functions such as sensory processing,
memory formation, reasoning, and motor control.
Much work has been put into neural simulators
that attempt to recreate neuroscientific
data in precise detail with the thought that
cognition will emerge by connecting
detailed neuron models according
to the statistics of biological synapses \citep{markram2006}.
However, cognition has not yet emerged
from data-driven large scale models,
and there are good reasons to think
that cognition may never emerge \citep{eliasmith2013a}.
At the other end of the spectrum,
cognitive architectures \citep{anderson2004,aisa2008}
and machine learning approaches \citep{hinton2006}
have solved high-dimensional statistical problems,
but do so without respecting biological constraints.

Nengo\footnote{Nengo is available for download through
  the Python Package Index at
  \url{https://pypi.python.org/pypi/nengo}.
  The full source is available at
  \url{https://github.com/ctn-waterloo/nengo}.}
is a neural simulator
based on a theoretical framework proposed
by \citet{eliasmith2003}
called the Neural Engineering Framework
(NEF).
The NEF is a large-scale modeling approach
that can leverage single neuron models
to build neural networks with
demonstrable cognitive abilities.
Nengo and the NEF has been used to build
increasingly sophisticated neural subsystems
for the last decade
(e.g., path integration [\citealp{conklin2005}],
working memory [\citealp{singh2006}],
list memory [\citealp{choo2010}],
inductive reasoning [\citealp{rasmussen2014}],
motor control [\citealp{dewolf2011}],
decision making [\citealp{stewart2012}])
culminating recently with Spaun,
currently the world's
largest functional brain model \citep{eliasmith2012}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight cognitive tasks
including memorizing lists and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed;
for example, sequentially presenting images
containing the characters $A3[1234]$ instructs Spaun
to memorize the list $1234$.
If asked to recall the memorized list,
Spaun generates motor commands for a simulated arm,
writing out the digits $1234$.
While the tasks that Spaun performs are diverse,
all of the tasks use a common set of
functional cortical and subcortical components.
Each functional component corresponds
to a brain area that has been hypothesized
to perform those functions
in the neuroscientific literature.

The NEF provides principles
to guide the construction of
a neural model
that incorporates anatomical constraints,
functional objectives, and
dynamical systems or control theory.
Constructing models from this starting point,
rather than from single cell electrophysiology
and connectivity statistics alone,
produces simulated data that explains
and predicts a wide variety of experimental results.
Single cell activity \citep{stewart2012},
response timing \citep{stewart2009a},
behavioral errors \citep{choo2010},
and age-related cognitive decline \citep{rasmussen2014},
of NEF-designed models match
physiological and psychological findings
without being built specifically into the design.
These results are a consequence of
the need to satisfy functional objectives
within anatomical and neurobiological constraints.

The transformation principle of the NEF
proposes that the connection weight matrix
between two neural populations
can compute a nonlinear function,
and can be factored into
two significantly smaller matrices.
By using these factors
instead of full connection weight matrices,
NEF-designed models are
more computationally efficient,
which allows Nengo to run
large-scale neural models
on low-cost commodity hardware.

In order to make Nengo more simple, extensible, and fast,
we have rewritten Nengo 2.0 from scratch in Python,
leveraging NumPy \citep{oliphant2007}
for manipulating large amounts of data.
While NumPy is its only dependency,
Nengo contains optional extensions
for plotting if Matplotlib is available
\citep{hunter2007}
and for interactive exploration
if IPython is available \citep{perez2007}.
Since Nengo only depends on one third-party library,
it is easy to integrate Nengo models
in arbitrary CPython programs,
opening up possibilities
for using neurally implemented algorithms
in web services, games, and other applications.

Nengo 2.0 has a simple object model,
which makes it easy to
document, test, and modify.
Model creation and simulation are decoupled,
allowing for models to be run with
other simulators as drop-in replacements for Nengo 2.0's
platform-independent reference simulator.
To date, we have implemented one other simulator
that uses PyOpenCL to take advantage
of a GPU or multicore CPU.
The OpenCL simulator can simulate
large models on the scale of Spaun
at least 50 times faster than Nengo 1.4
using inexpensive commodity hardware.

In all, Nengo 2.0 provides a platform for
simulating larger and more complex models than Spaun,
and can therefore further test the NEF
as a theory of neural computation.

\section{Neural Engineering Framework (NEF)} \label{sec:nef}

@@

In the encoding process, the input signal drives
each neuron based on its \textit{tuning curve},
which describes how likely
that neuron is to respond to a given input signal.
The tuning curve is a function of the gain
of a neuron (how quickly the activity rises),
the bias (the activity of a neuron given no signal),
and the \textit{encoding weight}
(the direction in the input vector space
that causes the neuron to be the most active).
Importantly, tuning curves can be determined
for any type of neuron,
and therefore the encoding process
(and the NEF as a whole)
is not dependent on any particular neuron model.
